---
title: "Final Project"
author: "Zhi Tu"
date: "2022-12-17"
output:
  pdf_document: default
  
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(shiny, leaflet, magrittr, dplyr, tidyverse, googleway, mapsapi, ggplot2, kableExtra, knitr)

```

## Data from GTFS

First, We get the data from the GTFS site: https://cdn.mbta.com/archive/archived_feeds.txt
After examining all these files, stop_times, stops, directions, and trip are most useful to use. For example, I can left_join trip onto the stop_times by the column `trip_id`. Also,  I can left_join directions onto the stop_times by the column `route_id` and `direction_id`. Since the file uploaded onto github cannot exceed 100mb, I concise all the files under the limits.

```{r, include=FALSE}
## Get stop and stoptimes from GTFS
# stoptime <- read.delim("MBTA_GTFS/stop_times.txt", sep = ",", header = TRUE)
# stoptime1 <- read.delim("stoptimes/stop_times.txt", sep = ",", header = TRUE)
# stoptime2 <- read.delim("stoptimes/stop_times_2.txt", sep = ",", header = TRUE)
# stoptime3 <- read.delim("stoptimes/stop_times_3.txt", sep = ",", header = TRUE)
# stoptime4 <- read.delim("stoptimes/stop_times_4.txt", sep = ",", header = TRUE)
# stoptime5 <- read.delim("stoptimes/stop_times_5.txt", sep = ",", header = TRUE)

# stoptime %<>% bind_rows(stoptime1)
# stoptime %<>% bind_rows(stoptime2)
# stoptime %<>% bind_rows(stoptime3)
# stoptime %<>% bind_rows(stoptime4)
# stoptime %<>% bind_rows(stoptime5)

stoptime <- read.delim("stop_times_final.txt", sep = ",", header = TRUE,fill = TRUE)
trip <- read.delim("trips.txt", sep = ",", header = TRUE)
direction <- read.delim("directions.txt", sep = ",", header = TRUE)
stop <- read.delim("stops.txt", sep = ",", header = TRUE,fill = TRUE)

stoptime %<>% select(1:5)
trip %<>% select(1:3,6)
stoptime %<>% left_join(trip, by="trip_id")
stoptime %<>% left_join(direction, by=c("route_id", "direction_id"))

stoptime$departure_time <- (strptime(stoptime$departure_time, format = "%H:%M:%S"))
stoptime %<>% select(-2,-7,-9,-10)
```

```{r, fig.cap="Stop times", echo=FALSE}
head(stoptime) 
```

On the above is the head rows of processed stoptime

```{r, include=FALSE}
# write.csv(stoptime,"stoptime_final.csv")
```


## Bus 

After getting all the stoptimes for different public transportation from GFTS, I would like to get stop informations.
For buses, I read the bus stop info from site: https://mbta-massdot.opendata.arcgis.com/maps/MassDOT::mbta-systemwide-gtfs-map/about.
After reading the file, I remove NAs in the Routes of buses and select columns with useful infos.

```{r, include=FALSE}
## Bus stops
bus_stop <- read.csv("bus stop.csv")
bus_stop %<>% filter(Routes != "" & Routes !="#N/A")
bus_stop %<>% select(1,2,4,6,28)
```

```{r, include=FALSE}
# write.csv(bus_stop,"bus_stop_final.csv")
```

Next, I find the processed bus arrival and departure times file on site:
https://mbta-massdot.opendata.arcgis.com/datasets/mbta-bus-arrival-departure-times-2021/about.


```{r, echo=FALSE, fig.cap="Bus Routes"}
## Bus departure time
# bus_1 <- read.csv("MBTA_Bus_Times_2021/MBTA-Bus-Arrival-Departure-Times_2021-01.csv")
# bus_route01 <- bus_1 %>% filter(route_id=="01")
bus_route01 <-  read.csv("route01_final.csv")
bus_route01 %<>% select(3,9,4,11,12)
head(bus_route01) 
```

As you can see, we need to convert the departure time on each stop to the time period between stops. Specifically, if the route has stop order from 1 to 9, then we need to have 8! combinations of time period to be calculated. 


Then, we can left join the departure stop info and the arrival stop info onto the combination to get all the time period from all the routes of buses.

```{r, include=FALSE}
# ## Calculate travel time based on departure time
# 
# bus_route01_from <-  bus_route01 %>% rename("from_stop_id"=stop_id)
# bus_route01_to <-  bus_route01 %>% rename("to_stop_id"=stop_id)
# 
# ## Get all combinations of stop orders
# comb <- bus_route01 %>% 
#   group_by(half_trip_id) %>% 
#   summarise(stop_id = combn(stop_id, 2, simplify = F)) %>% 
#   unnest_wider(stop_id, names_sep = "_")
# 
# ## Left join stop info onto stop orders
# comb %<>% rename("from_stop_id"=stop_id_1, "to_stop_id"=stop_id_2)
# bus_route01_from %<>% rename("start_time"=actual, "time_point_start"=time_point_order) %>% select(1:5,7,11)
# comb %<>% left_join(bus_route01_from, by=c("from_stop_id","half_trip_id"))
# bus_route01_to %<>%  rename("end_time"=actual, "time_point_end"=time_point_order) %>% select(4,5,7,11)
# comb %<>% left_join(bus_route01_to, by=c("to_stop_id","half_trip_id"))
# 
# ## Calculate time period in minutes
# comb$start_time <- as.character(comb$start_time)
# comb$end_time <- as.character(comb$end_time)
# comb %<>% filter(start_time!= "" & end_time!= "")
# route01 <- comb %>% mutate(time= difftime(end_time, start_time,units="mins"))
# route01$time <- as.numeric(route01$time)
# route01$time <- round(route01$time,2)
```



```{r, include=FALSE}
# write.csv(route01,"route01_final.csv")
```

\newpage

## Ferry

Next, we do the same thing for the ferry stop info. After getting the data, I found some names are named different across the file especially between the stop info and stoptimes. So I need to change one so that they can match to each other. 

```{r, include=FALSE}
## Ferry stops
ferry_stop <- read.csv("ferry stop.csv")
ferry_stop %<>% select(1:4, 6, 9, 13, 19, 14)
ferry_stop$stop_name[ferry_stop$stop_name=="Long Wharf (North)"] <- "Long Wharf N"
ferry_stop$stop_name[ferry_stop$stop_name=="Long Wharf (South)"] <- "Long Wharf S"
ferry_stop$stop_name[ferry_stop$stop_name=="Logan Airport Ferry Terminal"] <- "Logan"
ferry_stop %<>% slice(6) %>% mutate(stop_name="LOGAN") %>% bind_rows(ferry_stop)
ferry_stop$stop_name[ferry_stop$stop_name=="George's Island"] <- "Georges"
ferry_stop %>% select(1,2,4,5)
```

```{r, fig.cap="Ferry Stops"}
head(ferry_stop) 
```


```{r, include=FALSE}
# write.csv(ferry_stop,"ferry_stop_final.csv")
```

The data I got the daily info from:https://mbta-massdot.opendata.arcgis.com/search?tags=ferry
I then clean the data so that it can be match to the ferry stop info file. 

```{r, include=FALSE}
## Ferry time travels
# ferry_daily <- read.csv("MBTA_Ferry_Daily.csv")
ferry_daily <- read.csv("ferry_daily_final.csv")
ferry_stop_depart <- ferry_stop %>% mutate(departure_terminal= stop_name)

ferry_daily$actual_departure <- as.character(ferry_daily$actual_departure)
ferry_daily$actual_arrival <- as.character(ferry_daily$actual_arrival)
ferry_daily %<>% filter(actual_departure!= "" & actual_arrival!= "")
ferry_daily %<>% mutate(time= difftime(actual_arrival, actual_departure,units="mins"))
ferry_daily$time <- as.numeric(ferry_daily$time)
ferry_daily$time <- round(ferry_daily$time,2)
ferry_daily %<>% filter(departure_terminal != "Navy Yard" & arrival_terminal != "Navy Yard")
ferry_daily %<>% select(3,10,12, 17, 18, 23)
```

```{r, fig.cap="Ferry Travel Times"}
head(ferry_daily)
```

```{r, include=FALSE}
# write.csv(ferry_daily,"ferry_daily_final.csv")
```

\newpage 

## Commuter Rail

Commuter Rail is exactly same the buses, we need to calculate the time period based on the departure time.
Here is the results:

```{r, include=FALSE}
## Commuter rail stop info
rail_stop <- read.csv("rail stop.csv")
rail_stop %<>% select(4,6, 10, 11)
```

```{r, fig.cap="Commuter Rail Stops"}
head(rail_stop) 
```


```{r, include=FALSE}
# write.csv(rail_stop,"rail_stop_final.csv")
```

```{r, include=FALSE}
## Commuter rail time travels info
## Get commuter rail time travels info from stoptimes
rail_stoptime <- stoptime %>% filter(grepl( "CR-",route_id))

rail_from <-  rail_stoptime %>% rename("from_stop_id"=stop_id)
rail_to <-  rail_stoptime %>% rename("to_stop_id"=stop_id)

## Get all combinations of stop orders
comb_rail <- rail_stoptime %>% 
  group_by(trip_id) %>% 
  summarise(stop_id = combn(stop_id, 2, simplify = F)) %>% 
  unnest_wider(stop_id, names_sep = "_")

## Left join stop info onto stop orders
comb_rail %<>% rename("from_stop_id"=stop_id_1, "to_stop_id"=stop_id_2)
# rail_from %<>% rename("start_time"=departure_time, "time_point_start"=stop_sequence) %>% select(1,3:6,8,10)
rail_from %<>% rename("start_time"=departure_time, "time_point_start"=stop_sequence)
comb_rail %<>% left_join(rail_from, by=c("from_stop_id","trip_id"))
rail_to %<>%  rename("end_time"=departure_time, "time_point_end"=stop_sequence) %>% select(1:4)
comb_rail %<>% left_join(rail_to, by=c("to_stop_id","trip_id"))

## Calculate time period in minutes
# comb_rail$start_time <- as.character(comb_rail$start_time)
# comb_rail$end_time <- as.character(comb_rail$end_time)
# comb_rail %<>% filter(start_time!= "" & end_time!= "")
# comb_rail$start_time <- (strptime(comb_rail$start_time, format = "%H:%M:%S"))
# comb_rail$end_time <- (strptime(comb_rail$end_time, format = "%H:%M:%S"))
rail <- comb_rail %>% mutate(time= difftime(end_time, start_time,units="mins"))
rail$time <- as.numeric(rail$time)
rail$time <- round(rail$time,2)
rail %<>% select(6, 2, 4,3,8,10)
```

```{r, fig.cap="Commuter Rail Travel Times"}
head(rail) 
```


```{r, include=FALSE}
# write.csv(rail,"rail_final.csv")
```

\newpage

## MTBA data

Lastly, we got the MTBA data from the travel time file: https://mbta-massdot.opendata.arcgis.com/datasets/mbta-travel-times-2021/about


```{r, include=FALSE}
## Rapid Transit travel times
# tt_q1_2021 <- read.csv("Travel_Times_2021/HRTravelTimesQ1_21.csv")
# tt_q1_2021_lr <- read.csv("Travel_Times_2021/LRTravelTimesQ1_21.csv")

# tt_q4_2021 <- read.csv("Travel_Times_2021/HRTravelTimesQ4_21.csv")
# tt_q4_2021_lr <- read.csv("Travel_Times_2021/LRTravelTimesQ4_21.csv")
# dates_1 <- c("2021-10-01", "2021-10-04",
#              "2021-11-02", "2021-11-05", "2021-12-07")
# tt_q4_2021_lr %<>% filter(service_date %in% dates_1)
# tt_q4_2021 %<>% filter(service_date %in% dates_1)
# tt_q4_2021_lr %<>% bind_rows(tt_q4_2021)

# tt_q1_2022 <- read.csv("TravelTimes_2022/2022-Q1_HRTravelTimes.csv")
# tt_q1_2022_lr <- read.csv("TravelTimes_2022/2022-Q1_LRTravelTimes.csv")
# dates_2 <- c("2022-01-01", "2022-01-02","2022-01-03","2022-01-04","2022-01-07",
#              "2022-02-01", "2022-02-04","2022-02-05","2022-02-06","2022-02-07",
#              "2022-03-03", "2022-03-04","2022-03-05","2022-03-06","2022-03-07")
# tt_q1_2022_lr %<>% filter(service_date %in% dates_2)
# tt_q1_2022 %<>% filter(service_date %in% dates_2)
# tt_q1_2022_lr %<>% bind_rows(tt_q1_2022)


# tt_q2_2022 <- read.csv("TravelTimes_2022/2022-Q2_HRTravelTimes.csv")
# tt_q2_2022_lr <- read.csv("TravelTimes_2022/2022-Q2_LRTravelTimes.csv")
# dates_3 <- c("2022-04-02", "2022-04-03","2022-04-05","2022-04-06",
#              "2022-05-01", "2022-05-02","2022-05-03","2022-05-04",
#              "2022-06-01", "2022-06-05","2022-06-06","2022-06-07")
# tt_q2_2022_lr %<>% filter(service_date %in% dates_3)
# tt_q2_2022 %<>% filter(service_date %in% dates_3)
# tt_q2_2022_lr %<>% bind_rows(tt_q2_2022)
# 
# 
# tt_q3_2022 <- read.csv("TravelTimes_2022/2022-Q3_HRTravelTimes.csv")
# tt_q3_2022_lr <- read.csv("TravelTimes_2022/2022-Q3_LRTravelTimes.csv")
# dates_4 <- c("2022-07-04", "2022-07-05","2022-07-06","2022-07-07",
#              "2022-08-01", "2022-08-02","2022-08-04","2022-08-06",
#              "2022-09-01", "2022-09-05","2022-09-06","2022-09-07")
# tt_q3_2022_lr %<>% filter(service_date %in% dates_4)
# tt_q3_2022 %<>% filter(service_date %in% dates_4)
# tt_q3_2022_lr %<>% bind_rows(tt_q3_2022)
# 
# tt_q4_2021_lr %<>% bind_rows(tt_q1_2022_lr)
# tt_q4_2021_lr %<>% bind_rows(tt_q2_2022_lr)
# tt_q4_2021_lr %<>% bind_rows(tt_q3_2022_lr)
# tt_q1_2021_lr <- tt_q4_2021_lr

tt_q1_2021_lr <- read.csv("tt_q1_2021_lr_final.csv")
tt_q1_2021_lr %<>% select(5,3,7,4,8,9)
```

```{r, fig.cap="Rapid Transit Travel Times"}
head(tt_q1_2021_lr) 
```


```{r, include=FALSE}
# write.csv(tt_q1_2021_lr, "tt_q1_2021_lr_final.csv")
```


## Graph

I start the exploration on the MTBA data. At the beginning, I separate the subway by its `route_id` to different lines. 

```{r}
unique(tt_q1_2021_lr$route_id)
```


```{r, include=FALSE}
green_b <- tt_q1_2021_lr %>% filter(route_id=="Green-B")
green_c <- tt_q1_2021_lr %>% filter(route_id=="Green-C")
green_d <- tt_q1_2021_lr %>% filter(route_id=="Green-D")
green_e <- tt_q1_2021_lr %>% filter(route_id=="Green-E")
orange <- tt_q1_2021_lr %>% filter(route_id=="Orange")
blue <- tt_q1_2021_lr %>% filter(route_id=="Blue")
red <- tt_q1_2021_lr %>% filter(route_id=="Red")
```

Take the Green-B line as an example:

```{r, echo=FALSE}
head(green_b)
```

Here is the graph of time distribution between stops that I would like to include in my shinyApp:

```{r, echo=FALSE}
green_b %<>% filter(from_stop_id=="70111" & to_stop_id=="70107")
green_b %>% ggplot(aes(x=travel_time_sec)) +
  geom_histogram(colour="black", fill="white") + 
  ggtitle("Time Distribution graph") +
  labs(x= "Time in Seconds")
```

In addition to graph, I would also like a summary table for the time distribution between specific stops.

```{r}
summary(green_b$travel_time_sec)
```







